---
title: "Demographic joins"
format: html
editor: visual
---

## Demographics

```{r}
library(dplyr)
library(ggplot2)
# install.packages("sf") # https://github.com/r-spatial/sf/issues/2035
library(sf)
library(tidycensus)
options(tigris_use_cache = TRUE)
library(tidyr)
library(purrr)

```


```{r}
bg_shapes <- get_acs(
  geography = "block", 
  variables = c("B02012_001", "B02001_001", "B02001_002", "B02009_001", "B02010_001", 
                "B02011_001", "B02012_001", "B19013_001", "B03003_001", "B03003_003"),
  state = "CA", 
  year = 2020,
  cache_table = TRUE,
  geometry = TRUE
) 



```

```{r}
block_shapes <- get_decennial(
  geography = "block", 
  variables = c("P3_002N", "P3_003N", "P3_004N", "P3_005N", "P3_006N", 
                "P3_007N", "P3_008N", "P4_001N", "P4_002N"),
  state = "CA", 
  year = 2020,
  cache_table = TRUE,
  geometry = TRUE
) 
```

Note: 2020 decennial Census data use differential privacy, a technique that introduces errors into data to preserve respondent confidentiality. ℹ Small counts should be interpreted with caution. ℹ See https://www.census.gov/library/fact-sheets/2021/protecting-the-confidentiality-of-the-2020-census-redistricting-data.html for additional guidance.

Total P3_002N\
White P3_003N\
Black P3_004N\
Indigenous P3_005N\
Asian P3_006N\
Paci_Isl P3_007N\
Other P3_008N Total_hisp P4_001N\
Hisp_Lat P4_002N

```{r}
block_shapes_wide <- block_shapes %>% 
  pivot_wider(names_from = "variable", values_from = "value") %>% 
  rename(Total = P3_002N,	
        White = P3_003N,	
        Black = P3_004N,
        Indigenous = P3_005N,	
        Asian = P3_006N,
        Paci_Isl = P3_007N,	
        Other = P3_008N,
        Total_Hisp = P4_001N,	
        Hisp_Lat = P4_002N) %>% 
  mutate(White = White / Total,
         Black = Black / Total,
         Indigenous = Indigenous / Total,
         Asian = Asian / Total,
         Paci_Isl = Paci_Isl / Total,
         Other = Other / Total,
         Hisp_Lat = Hisp_Lat / Total_Hisp
         )

block_shapes_tall_props <- block_shapes_wide %>% 
  pivot_longer(c(White, Black, Asian, Paci_Isl, Indigenous, Other, Hisp_Lat), 
               names_to = "race", values_to = "prop")

ggplot(block_shapes_tall2, aes(x = race, y = prop, color = race)) + geom_col()

block_shapes_vals <- block_shapes %>% 
  filter(!(Total == 0 & Total_Hisp == 0))
```

```{r}
ggplot(block_shapes_vals, aes(Total)) + geom_histogram() + scale_x_log10()
```

Vars

Race total: B02001_001 White alone B02001_002 B02009_001 Black alone or in combination B02010_001 Native American alone or in combination B02011_001 Asian alone or in combination B02012_001 Native Hawaiian or Pacific Islander alone or in combination B19013_001 median household income Hispanic total: B03003_001 B03003_003 yes hispanic or latino

```{r}
bg_shapes <- st_transform(bg_shapes, 4269) # make types compatible
watershapes <- st_transform(watershapes, 4269)

join <- watershapes %>% 
  inner_join(UCMR_CA, by = c("pwsid" = "PWSID")) %>% 
  select(pwsid, PWSName, primacy_agency_code, primary_source_code, tier, FacilityWaterType, 
         CollectionDate, Contaminant, MRL, AnalyticalResultsSign, AnalyticalResultValue.µg.L.) %>% 
  st_intersection(bg_shapes) %>% 
  mutate(pwi_county = paste0(pwsid, "-", GEOID))


joinTo <- join %>% 
  #--- get area ---#
  mutate(area = as.numeric(st_area(.))) %>% 
  #--- get area-weight by HUC unit ---#
  group_by(pwsid) %>% 
  mutate(weight = area / sum(area)) %>% 
  #--- calculate area-weighted corn acreage by HUC unit ---#
  summarize(aw_acres = sum(weight * acres)) 

```

Create bool for yes, all present Create categorical for "predominantly \_\_\_ population"
