---
title: "Comparing PWS shape data"
format: html
editor: visual
---

## PWS Shape Data

```{r}
library(dplyr)
library(ggplot2)
# install.packages("sf") # https://github.com/r-spatial/sf/issues/2035
library(sf)
library(tidycensus)
options(tigris_use_cache = TRUE)
library(tidyr)
library(purrr)
```

```{r}
SLwatershapes <- st_read("data/temm.gpkg") # https://community.rstudio.com/t/how-to-open-a-geopackage-file/41027

```

```{r}
govwatershapes <- st_read("data/SABL_Public_230403") %>% 
  st_transform(4269) %>% 
  rename("PWSID" = "SABL_PWSID")

```

```{r}

ggplot(data = watershapes) +
  geom_sf() +
  scale_x_continuous(limits = c(-114, -125)) + 
  scale_y_continuous(limits = c(32, 43))


ggplot(data = govwatershapes %>% filter(VERIFIED_S == "Verified")) +
  geom_sf()

govwatershapes %>% 
  st_drop_geometry() %>% 
  group_by(VERIFIED_S) %>% 
  summarise(Number = n())
watershapes %>%
  st_drop_geometry() %>% 
  group_by(tier) %>% 
  summarise(Number = n())
```

Next step: find data of indigenous people's land and join to govwatershapes

-   Write about it in methods

Next next step: wrangling demographic data. Reference methods section for this

You got this, Audrey!!!

```{r}


# CAtest <- watershapes %>% 
#   filter(pwsid %in% unique(UCMR_CA_indig$PWSID)) %>% 
#   rename("PWSID" = "pwsid",
#          "WATER_SY_1" = "pws_name",
#          "geometry" = "geom") %>% 
#   select(PWSID, WATER_SY_1, geometry) %>% 
#   st_transform(4269)



SLwatershapes %>% 
  filter(UCMR_CA_indig)

```

```{r}
# Joining UCMR data with 
UCMRshapes_inner <- CAwatershapes %>% 
  inner_join(UCMR_CA_indig) %>% 
  select(PWSID, PWSName, FacilityWaterType, Contaminant, 
         AnalyticalResultsSign, AnalyticalResultValue.µg.L., State, geometry) %>% 
  rename("ContamLevel" = "AnalyticalResultValue.µg.L.")

UCMRshapes_left <- UCMR_CA_indig %>% 
  left_join(CAwatershapes) %>% 
  select(PWSID, PWSName, FacilityWaterType, Contaminant, 
         AnalyticalResultsSign, AnalyticalResultValue.µg.L., State, geometry) %>% 
  rename("ContamLevel" = "AnalyticalResultValue.µg.L.")

setdiff(UCMRshapes_left %>% select(PWSID) %>% st_drop_geometry() %>% unique() %>% pull(), 
        UCMRshapes_inner %>% select(PWSID) %>% st_drop_geometry() %>% unique() %>% pull())

UCMRshapes_left %>% 
  st_is_empty()

ggplot(UCMRshapes_inner %>% select(geometry) %>% unique()) +
  geom_sf()

UCMR_CA_indig %>% select(PWSID) %>% unique()
UCMRshapes_left %>% select(PWSID) %>% st_drop_geometry() %>% unique() %>% pull() %>% length()

```

We henceforth use UCMRshapes_inner

Now I think we're joining UCMRshapes_inner with demographic data

First we load demographic data\\

\*record scratch\* I think since there are so many rows from contaminants, I should join by water shape and demographics first and THEN join those to contaminants. At least now I know about the contaminants deal.

```{r}


```
